Real-space computation of $E$/$B$-mode maps I: Formalism and Compact Kernels

### 2017-11-24 articles/formalism/sections/real_space_formalism.tex
The spatial argument of spin-0 construction in section 3.1 is now more solid. I am satisfied with this explanation now. Still not happy with the explanation for convolution and radiating interpretations of the real space kernels.

Using \eq{eq:ebdef} and the properties of the spin raising and lowering operators given in \eq{eq:spinopylm} it can be shown that the scalar fields $\mathcal{E}/\mathcal{B}$ are defined via the following set of equations,
%
\beq \label{eq:pseudo}
\mathcal{E}(\hat{n}) = \sum_{\ell m} a^{E}_{\ell m} \sqrt{\frac{(\ell+2)!}{(\ell-2)!}} Y_{\ell m} (\hat{n}) ~\,;~ \mathcal{B}(\hat{n})  =\sum_{\ell m} a^{B}_{\ell m} \sqrt{\frac{(\ell+2)!}{(\ell-2)!}} Y_{\ell m} (\hat{n}) \,,
\eeq
%
where the harmonic coefficients of  $\mathcal{E}/\mathcal{B}$ fields are related to the harmonic coefficients of the spin-2 polarization field via the following equations,
%
\beq\label{eq:x2eb}
a^{E}_{\ell m} = -\frac{1}{2} \Big[ {}_{+2}\tilde{X}_{\ell m} + _{-2}\tilde{X}_{\ell m} \Big] ~\,;~a^{B}_{\ell m} = -\frac{1}{2i} \Big[ {}_{+2}\tilde{X}_{\ell m} - _{-2}\tilde{X}_{\ell m} \Big] 
\eeq
%
In the remainder of this article, we will work with the scalar $E$ and pseudo scalar $B$ fields as defined by the following expressions, 
%
\beq \label{eq:realeb}
E(\hat{n}) = \sum_{\ell m} a^{E}_{\ell m} Y_{\ell m} (\hat{n}) ~\,;~ B(\hat{n})  =\sum_{\ell m} a^{B}_{\ell m} Y_{\ell m} (\hat{n}) \,.
\eeq
%
Note that the $[E,B]$ fields are merely filtered versions of the fields $[\mathcal{E},\mathcal{B}]$, as their spherical harmonic coefficients of expansion differ by the factor $\sqrt{\frac{(\ell+2)!}{(\ell-2)!}}$. %We make this choice since the CMB spectra are more closely related to the fields E \& B.
%--------------------------------------------------------
%--------------------------------------------------------
\subsection{Matrix notation} \label{sec:mat_pol_intro}
In this section we cast the relations introduced in Sec.~\ref{sec:pol-primer} in matrix notation\footnote{While we work with the matrix and vector sizes given in terms of some pixelization parameter $\rm N_{\rm pix}$, all the relations are equally valid in the continuum limit attained by allowing $\rm N_{\rm pix}\rightarrow \infty$}. This representation will make transparent the derivation of the real space operators we discuss in the following sections. We adopt a convention in which real space quantities are denoted by bar-ed variable while those in harmonic space are denoted by tilde-ed variables.

@@

where the symbol ${}_{0}Y_e$ is used to denote the sub-matrix ${}_{0}Y_{\hat{n}_e \times \ell m} \equiv {}_{0}Y_{\ell m}(\hat{n}_e)$, the symbol ${}_{\pm 2}Y^{T*}_q$ is used to denote the matrix ${}_{\pm 2}Y^*_{\ell m \times \hat{n}_q} \equiv {}_{\pm 2}Y^*_{\ell m}(\hat{n}_q)$ and the summation is over the multipole indices $\ell,m$. \revisit{We purposefully use the index ``e'' to denote the location where the scalar fields are measured and the index ``q'' to denote the location where the Stokes parameters are measured.} Using the conjugation properties of the spin spherical harmonic functions it can be shown that the following relation holds true,

@@

\textit{Radiating:} On the contrary, while working from the frame of the pixel ``q'', the Euler angles to the surrounding pixels are: $\alpha_{qe}, \beta_{qe},\gamma_{qe}$. The E \& B mode maps resulting from the Stokes Q \& U parameters specified at location $\hat{n}_q$ is given by the following expression,
%
\beq  \label{eq:qu2eb_radiation_explicit}
\bmat E_e \\ B_e  \emat_{q} =- {{}_{\mm}f}(\beta_{qe},\ell_{\rm min},\ell_{\rm max})\bmat \cos(2 \alpha_{q e}) & \sin(2\alpha_{q e})\\  -\sin(2 \alpha_{q e})  & \cos(2 \alpha_{q e}) \emat  \bmat Q_{q} \\ U_{q}  \emat \Delta \Omega \,.
\eeq
%
The total map of E \& B modes can be simply evaluated by summing over the contribution from the Stokes parameters at each location $\hat{n}_q$ : $\bar{S} = \sum_{q=1}^{N_{\rm pix}} \bar{S}_q$. The above expression can again be cast in a more concise form as follows,
%
\begin{subequations} \label{eq:qu2eb_radiation_concise}
\beqry 
\left[E + iB\right] &=& -\Delta \Omega  \sum_{q=1}^{N_{\rm pix}} \Big[ {}_{+2}X(\hat{n}_{q}) e^{-i2\alpha_{q e}} \Big]  {{}_{\mm}f}(\beta_{q e}) \,, \\
&=&    \sum_{q=1}^{N_{\rm pix}} \Bigg\lbrace {}_{+2}X(\hat{n}_{q}) \cdot \left[ - \Delta \Omega \sum_{\ell=\ell_{\rm min}}^{\ell_{\rm max}} \sqrt{\frac{2 \ell+1}{4 \pi}}Y^*_{\ell 2}(\beta_{qe},\alpha_{qe}) \right] \Bigg\rbrace \,, \\
&=& \sum_{q=1}^{N_{\rm pix}} {}_{+2}X(\hat{n}_{q}) \cdot  \mathcal{M}_{Green}(\hat{n}_{q}) \,,
\eeqry
\end{subequations}
%
where ``$\cdot$'' denotes a simple scalar multiplication. \revisit{$\mathcal{M}_{Green}(\hat{n}_{q})$ is response of the real space operator $\bar{O}$ when operated on a Stokes field: $\delta(\hat{n}-\hat{n}_q) + i \delta(\hat{n}-\hat{n}_q)$, hence it can be thought of as the Green's function of the real space operator.} \comment{Is this interpretation correct ? Is the complex delta function a spin 2 function ?}

Recall that the product of two functions with spins $s_1$ and $s_2$ results in a function with spin $s_1 + s_2$: ${}_{s_1 +s_2}f = {}_{s_1}g \,{}_{s_2}h$ \comment{The following argument is that of the product of a number with a field, so is this relevant ? }. If one rotates by an angle $\phi$ about the local $\hat{z}_q$ axis, the local Stokes parameters transform as: ${}_{+2}X(\hat{n}_q) \xrightarrow{\mathcal{R}_{\hat{z}_q}(\phi)} {}_{+2}X(\hat{n}_q) e^{i2\phi} $. Since this angle $\phi$ can be thought of as the phase angle $\gamma$ in the $z-y_1-z_2$ convention of Euler angles, one can see that: $e^{-i2\alpha_{qe}} \xrightarrow{\mathcal{R}_{\hat{z}_q}(\phi)} e^{-i2\alpha_{qe}} e^{-i2\phi}$. The Euler angles $|\beta_{qe}|$ which measure the angular distance between pixels remains unaltered under this local rotation operation.  Given these transformation properties one can now note that the product ${}_{+2}X(\hat{n}_q)e^{-i2\alpha_{qe}}$ is invariant under rotations  $\mathcal{R}_{\hat{z}_q}(\phi)$ and hence must form a spin-0 field.  This argument makes intuitive, the construction of the spin-0 E and B modes of polarization from the measured Stokes parameters. \revisit{Here it is also interesting to note that the E-modes are constructed by product of functions $(U\sin{2 \alpha}, Q\cos{2\alpha})$ which have the same parity and hence have even parity while the B-modes are constructed by multiplying functions $(Q\sin{2 \alpha}, U\cos{2\alpha})$ of opposite parity and hence have an odd parity.}

Using \eq{eq:qu2eb_radiation_explicit} one can see that when $q=e$: $E_q=-{{}_{\mm}f}(\beta_{qq})Q_q$ and $B_q = -{{}_{\mm}f}(\beta_{qq})U_q$ which does not transform as a spin-0 field under local rotations unless ${{}_{\mm}f}(\beta_{qq}=0)=0$. One can make a similar argument when pixels ``q'' and ``e'' are diametrically opposite. Therefore the radial kernels have to satisfy the constraint that they vanish at $\beta = 0, \pi$. The radial kernel ${{}_{\mm}f}(\beta)$ is constructed by forming a weighted linear combination of $P_{\ell}^2$ Legendre polynomials which have the following limiting form: $P_{\ell}^2(\beta) \propto \sin^2{\beta} \rightarrow 0 $ as $\beta \rightarrow 0 ,\pi$ and therefore satisfy this necessary constraint.

The real space operator has a azimuthal part which depends only on the Euler angle $\alpha_{qe} (\gamma_{eq})$, has no multipole $\ell$ dependence and is the crucial operation which translates between the two different spin representation of CMB polarization. The radial part of the kernel is specified by the function ${}_{\mm}f(\beta,\ell_{\rm min},\ell_{\rm max})$, it depends only on the angular separation $|\beta_{eq}|$ between pixels and completely incorporates the multipole $\ell$ dependence of the kernel. It's the radial part of the kernel that determines the locality of the real space operator.
\subsection{Evaluating Stokes parameters $Q$ \& $U$ from scalar fields $E$ \& $B$}\label{sec:eb2qu}
The real space operator which translates $E$ \& $B$ fields to Stokes parameters $Q$ \& $U$ can be derived using a similar procedure. The inverse operator is given by the following expression,
\bar{P} &=& \bar{T}^{-1} *{{}_2\mathcal{Y}} *\tilde T *{{}_0\mathcal{Y}^{\dagger}}\bar{S} = \frac{1}{2} \bar{T}^{\dagger} *{{}_2\mathcal{Y}} *\tilde T *{{}_0\mathcal{Y}^{\dagger}}\bar{S} \,,  \\
&=&  \bar O^{-1} *\bar{S}\,.
We do not provide the explicit derivation here, since the procedure is identical to that discussed in the previous section. The inverse operator is given by the following expression,

>>>>

Using \eq{eq:ebdef} and the properties of the spin raising and lowering operators given in \eq{eq:spinopylm} it can be shown that the scalar fields $\mathcal{E}/\mathcal{B}$ are defined via the following set of equations,
%
\beq \label{eq:pseudo}
\mathcal{E}(\hat{n}) = \sum_{\ell m} a^{E}_{\ell m} \sqrt{\frac{(\ell+2)!}{(\ell-2)!}} Y_{\ell m} (\hat{n}) ~\,;~ \mathcal{B}(\hat{n})  =\sum_{\ell m} a^{B}_{\ell m} \sqrt{\frac{(\ell+2)!}{(\ell-2)!}} Y_{\ell m} (\hat{n}) \,,
\eeq
%
where the harmonic coefficients of  $\mathcal{E}/\mathcal{B}$ fields are related to the harmonic coefficients of the spin-2 polarization field via the following equations,
%
\beq\label{eq:x2eb}
a^{E}_{\ell m} = -\frac{1}{2} \Big[ {}_{+2}\tilde{X}_{\ell m} + {}_{-2}\tilde{X}_{\ell m} \Big] ~\,;~a^{B}_{\ell m} = -\frac{1}{2i} \Big[ {}_{+2}\tilde{X}_{\ell m} - {}_{-2}\tilde{X}_{\ell m} \Big] 
\eeq
%
In the remainder of this article, we will work with the scalar $E$ and pseudo scalar $B$ fields as defined by the following expressions, 
%
\beq \label{eq:realeb}
E(\hat{n}) = \sum_{\ell m} a^{E}_{\ell m} Y_{\ell m} (\hat{n}) ~\,;~ B(\hat{n})  =\sum_{\ell m} a^{B}_{\ell m} Y_{\ell m} (\hat{n}) \,.
\eeq
%
Note that the $[E,B]$ fields are merely filtered versions of the fields $[\mathcal{E},\mathcal{B}]$, as their spherical harmonic coefficients of expansion differ by the factor $\sqrt{\frac{(\ell+2)!}{(\ell-2)!}}$. %We make this choice since the CMB spectra are more closely related to the fields E \& B.
%--------------------------------------------------------
%--------------------------------------------------------
\subsection{Matrix notation} \label{sec:mat_pol_intro}
In this section we cast the relations introduced in Sec.~\ref{sec:pol-primer} in matrix notation\footnote{While we work with the matrix and vector sizes given in terms of some pixelization parameter $\rm N_{\rm pix}$, all the relations are equally valid in the continuum limit attained by allowing $\rm N_{\rm pix}\rightarrow \infty$}. This representation will make transparent the derivation of the real space operators we discuss in the following sections. We adopt a convention in which real space quantities are denoted by bar-ed variable while those in harmonic space are denoted by tilde-ed variables.

@@

where the symbol ${}_{0}Y_e$ is used to denote the sub-matrix ${}_{0}Y_{\hat{n}_e \times \ell m} \equiv {}_{0}Y_{\ell m}(\hat{n}_e)$, the symbol ${}_{\pm 2}Y^{T*}_q$ is used to denote the matrix ${}_{\pm 2}Y^*_{\ell m \times \hat{n}_q} \equiv {}_{\pm 2}Y^*_{\ell m}(\hat{n}_q)$ and the summation is over the multipole indices $\ell,m$. \revisit{We purposefully use the index ``e'' to denote the location where the scalar fields are measured and the index ``q'' to denote the location where the Stokes parameters are measured.} Using the conjugation properties of the spin spherical harmonic functions it can be shown that the following relation holds true, \comment{These constructions are dyadics}

@@

\textit{Radiating:} On the contrary, while working from the frame of the pixel ``q'', the Euler angles to the surrounding pixels are: $\alpha_{qe}, \beta_{qe},\gamma_{qe}$. The E \& B mode maps resulting from the Stokes Q \& U parameters specified at the location $\hat{n}_q$ is given by the following expression,
%
\beq  \label{eq:qu2eb_radiation_explicit}
\bmat E_e \\ B_e  \emat_{q} =- {{}_{\mm}f}(\beta_{qe},\ell_{\rm min},\ell_{\rm max})\bmat \cos(2 \alpha_{q e}) & \sin(2\alpha_{q e})\\  -\sin(2 \alpha_{q e})  & \cos(2 \alpha_{q e}) \emat  \bmat Q_{q} \\ U_{q}  \emat \Delta \Omega \,.
\eeq
%
The total map of E \& B modes can be simply evaluated by summing over the contribution from the Stokes parameters at each location $\hat{n}_q$ : $\bar{S} = \sum_{q=1}^{N_{\rm pix}} \bar{S}_q$. The above expression can again be cast in a more concise form as follows,
%
\begin{subequations} \label{eq:qu2eb_radiation_concise}
\beqry 
\left[E + iB\right] &=& -\Delta \Omega  \sum_{q=1}^{N_{\rm pix}} \Big[ {}_{+2}X(\hat{n}_{q}) e^{-i2\alpha_{q e}} \Big]  {{}_{\mm}f}(\beta_{q e}) \,, \\
&=&    \sum_{q=1}^{N_{\rm pix}} \Bigg\lbrace {}_{+2}X(\hat{n}_{q}) \cdot \left[ - \Delta \Omega \sum_{\ell=\ell_{\rm min}}^{\ell_{\rm max}} \sqrt{\frac{2 \ell+1}{4 \pi}}Y^*_{\ell 2}(\beta_{qe},\alpha_{qe}) \right] \Bigg\rbrace \,, \\
&=& \sum_{q=1}^{N_{\rm pix}} {}_{+2}X(\hat{n}_{q}) \cdot  \mathcal{M}_{Green}(\hat{n}_{q}) \,,
\eeqry
\end{subequations}
%
where ``$\cdot$'' denotes a simple scalar multiplication. \revisit{$\mathcal{M}_{Green}(\hat{n}_{q})$ is response of the real space operator $\bar{O}$ when operated on a Stokes field: $\delta(\hat{n}-\hat{n}_q) + i \delta(\hat{n}-\hat{n}_q)$, hence it can be thought of as the Green's function of the real space operator.} \comment{Is this interpretation correct ? Is the complex delta function a spin 2 function ?}

Recall that the product of two functions with spins $s_1$ and $s_2$ results in a function with spin $s_1 + s_2$: ${}_{s_1 +s_2}f = {}_{s_1}g \,{}_{s_2}h$ \comment{The following argument is that of the product of a number with a field, so is this relevant ?}. If one rotates the cartesian coordinates in the tangent plane at location $\hat{n}_q$ by an angle $\phi$ about the local $\hat{z}_q$ axis, the Stokes parameters in the new rotated frame are given by: ${}_{+2}X(\hat{n}_q) \xrightarrow{\mathcal{R}_{\hat{z}_q}(\phi)} {}_{+2}X(\hat{n}_q) e^{-i2\phi} $. \revisit{Recall that the Euler angle $\alpha_{qe}$ in the $z-y_1-z_2$ sense of rotations is such that it aligns the local $\hat{x}_q$-axis along the geodesic in the direction of $\hat{n}_e$. Now if the local planar axes are already rotated by an angle $\phi$ then $\alpha_{qe} \rightarrow \alpha_{qe} - \phi$ and therefore one can see that: $e^{-i2\alpha_{qe}} \xrightarrow{\mathcal{R}_{\hat{z}_q}(\phi)} e^{-i2\alpha_{qe}} e^{i2\phi}$.} The Euler angles $|\beta_{qe}|$ which measure the angular distance between pixels remains unaltered under this local rotation operation.  Given these transformation properties one can now note that the product ${}_{+2}X(\hat{n}_q)e^{-i2\alpha_{qe}}$ is invariant under rotations  $\mathcal{R}_{\hat{z}_q}(\phi)$ and hence must form a spin-0 field.  This argument makes intuitive, the construction of the spin-0 E and B modes of polarization from the measured Stokes parameters. \revisit{Here it is also interesting to note that the E-modes are constructed by product of functions $(U\sin{2 \alpha}, Q\cos{2\alpha})$ which have the same parity and hence have even parity while the B-modes are constructed by multiplying functions $(Q\sin{2 \alpha}, U\cos{2\alpha})$ of opposite parity and hence have an odd parity.}

Using \eq{eq:qu2eb_radiation_explicit} one can see that when $q=e ~(\beta_{qq}=0\,,\, \alpha_{qq}=0,2\pi,4\pi,...)$: $E_q=-{{}_{\mm}f}(\beta_{qq})Q_q$ and $B_q = -{{}_{\mm}f}(\beta_{qq})U_q$ which does not transform as a spin-0 field under local rotations unless ${{}_{\mm}f}(\beta_{qq}=0)=0$. One can make a similar argument when pixels ``q'' and ``e'' are diametrically opposite. Therefore the radial kernels have to satisfy the constraint that they vanish at $\beta = 0, \pi$. The radial kernel ${{}_{\mm}f}(\beta)$ is constructed by forming a weighted linear combination of $P_{\ell}^2$ Legendre polynomials which have the following limiting form: $P_{\ell}^2(\beta) \propto \sin^2{\beta} \rightarrow 0 $ as $\beta \rightarrow 0 ,\pi$ and therefore satisfy this necessary constraint.

The real space operator has a azimuthal part which depends only on the Euler angle $\alpha_{qe} (\gamma_{eq})$, has no multipole $\ell$ dependence and is the crucial operation which translates between the two different spin representation of CMB polarization. The radial part of the kernel is specified by the function ${}_{\mm}f(\beta,\ell_{\rm min},\ell_{\rm max})$, it depends only on the angular separation $|\beta_{eq}|$ between pixels and completely incorporates the multipole $\ell$ dependence of the kernel. It's the radial part of the kernel that determines the locality of the real space operator.
%--------------------------------------------------------
%--------------------------------------------------------
\subsection{Evaluating Stokes parameters $Q$ \& $U$ from scalar fields $E$ \& $B$}\label{sec:eb2qu}
The real space operator which translates $E$ \& $B$ fields to Stokes parameters $Q$ \& $U$ can be derived using a similar procedure. The inverse operator is given by the following expression,
%
\begin{subequations}
\beqry
\bar{P} &=& \bar{T}^{-1} *{{}_2\mathcal{Y}} *\tilde T *{{}_0\mathcal{Y}^{\dagger}}\bar{S} = \frac{1}{2} \bar{T}^{\dagger} *{{}_2\mathcal{Y}} *\tilde T *{{}_0\mathcal{Y}^{\dagger}}\bar{S} \,,  \\
&=&  \bar O^{-1} *\bar{S}\,.
\eeqry
\end{subequations}
%
Providing the explicit derivation here will be redundant with that described in the previous sections and hence we avoid that. The inverse operator $\bar{O}^{-1}$ can be derived by realizing that $\mm^{-1}$ is given by the following expression,
%
\beq
\mm^{-1} = \sum_{\ell m}  {{}_{2}}Y_{\ell m}(\hat n_q) {}_{0}Y^*_{\ell m}(\hat n_e) \, 
\eeq
%

 The inverse operator is given by the following expression,
